% R bootcamp, Module 3: Working with objects and data
% August 2021, UC Berkeley
% Chris Paciorek

```{r chunksetup, include=FALSE} 
# include any code here you don't want to show up in the document,
# e.g. package and dataset loading
library(gapminder)
```

# Matrices and Arrays

If you need to do numeric calculations on an entire non-vector object (dimension > 1), you generally want to use matrices and arrays, not data frames.

- Matrices are two-dimensional collections of values of the same type
- We can have numeric, integer, character, or logical matrices, character matrices.
- You can't mix types within a matrix

```{r}
mat <- matrix(rnorm(12), nrow = 3, ncol = 4)
mat

# vectorized calcs work with matrices too
mat*4
mat <- cbind(mat, 1:3)
mat

# Let's convert the gapminder dataframe to a matrix:
gm_mat <- as.matrix(gapminder[ , c('lifeExp', 'gdpPercap')])
head(gm_mat)

# What do you think will happen?
gm_mat <- as.matrix(gapminder)
head(gm_mat)
```

Arrays are like matrices but can have more or fewer than two dimensions.
```{r}
arr <- array(rnorm(12), c(2, 3, 4))
arr
```

# Matrices: quick quiz

**POLL 3A**: Recall the gap dataframe has columns that are numeric and columns that are character strings. What do you think will happen if we do this:

```
as.matrix(gapminder)
```

1) it will convert to a matrix with no changes
2) all numeric columns will be converted to character strings
3) R will throw an error
4) all character columns will be converted to numeric values
5) R will drop some of the columns

# Attributes

Objects have *attributes*.

```{r attr}
attributes(mat)
rownames(mat) <- c('first', 'middle', 'last')
mat
attributes(mat)

names(attributes(gapminder))
attributes(gapminder)$names
attributes(gapminder)$row.names[1:10]
```

Now let's do a bit of manipulation and see if you can infer how R represents matrices internally.

# Attributes: quick quiz

**POLL 3B**: Consider our matrix 'mat':

```
mat <- matrix(1:16, nrow = 4, ncol = 4)
     [,1] [,2] [,3] [,4]
[1,]    1    5    9   13
[2,]    2    6   10   14
[3,]    3    7   11   15
[4,]    4    8   12   16
```

Suppose I run this code: mat[4]

What do you think will be returned?

1) 13
2) 4
3) 13, 14, 15, 16
4) 4, 8, 12, 16
5) an error


```{r what_is_a_matrix, eval = FALSE}
mat[4]
attributes(mat) <- NULL
mat
is.matrix(mat)
```

**Question**: What can you infer about what a matrix is in R?

**Question**: What kind of object are the attributes themselves? How do I check?

# Matrices are stored column-major

This is like Fortran, MATLAB and Julia but not like C or Python(numpy). 

```{r}
mat <- matrix(1:12, 3, 4)
mat
c(mat)
```
You can go smoothly back and forth between a matrix (or an array) and a vector:
```{r}
identical(mat, matrix(c(mat), 3, 4))
identical(mat, matrix(c(mat), 3, 4, byrow = TRUE))
```

This is a common cause of bugs!


# Missing values and other special values

Since it was designed by statisticians, R handles missing values very well relative to other languages.

* `NA` is a missing value
```{r, fig.cap = ""}
vec <- rnorm(12)
vec[c(3, 5)] <- NA
vec
length(vec)
sum(vec)
sum(vec, na.rm = TRUE)
hist(vec)
is.na(vec)
```
Be careful because many R functions won't warn you that they are ignoring the missing values.

* To infinity and beyond
```{r}
big <- 1e500 
big
big + 7
```

* `NaN` stands for Not a Number
```{r}
sqrt(-5)
big - big
1/0
```

* `NULL`
```{r nulldata}
vec <- c(vec, NULL) 
vec
length(vec)
a <- NULL
a + 7
a[3, 4]
is.null(a)
myList <- list(a = 7, b = 5)
myList$a <- NULL  # works for data frames too
myList
```

`NA` can hold a place but `NULL` cannot.
`NULL` is useful for having a function argument default to 'nothing'. See `help(crossprod)`, which can compute either $X^{\top}X$ or $X^{\top}Y$.  

# Missing values: quick quiz

**POLL 3C**

*Question 1*: Consider the following vector:

```
vec <- c(3, NA, 7)
```

What is vec[2]:

1) NA
2) 7

*Question 2*: Consider this vector:

```
vec <- c(3, NULL, 7)
```

What is vec[2]:

1) NULL
2) NA
3) 7
4) 3

*Question 3*: Consider this list:

```
mylist <- list(3, NULL, 7)
```

What is mylist[[2]]:

1) 7
2) NULL
3) NA
4) 3

*Question 4*: Consider this code:

```
mylist <- list(3, 5, 7)
mylist[[2]] <- NULL
```

What is length(mylist):

1) 3
2) 2
3) 1


# Logical vectors
```{r}
gapminder2007 <- gapminder[gapminder$year == 2007, ]

wealthy <- gapminder2007$gdpPercap > 35000
healthy <- gapminder2007$lifeExp > 75

head(wealthy)
table(wealthy)

# note the vectorized boolean arithmetic
gapminder2007[wealthy & healthy, ]
gapminder2007[wealthy | healthy, ]
gapminder2007[wealthy & !healthy, ]

# what am I doing here?
sum(healthy)
mean(healthy)
```

**Question**: What do you think R is doing to do arithmetic on logical vectors?


# Data frames

A review from Module 1...

- Data frames are combinations of vectors of the same length, but can be of different types
- Data frames are what is used for standard rectangular (record by field) datasets, similar to a spreadsheet
- Data frames are a functionality that both sets R aside from some languages (e.g., Matlab) and provides functionality similar to some statistical packages (e.g., Stata, SAS)

```{r}
class(gapminder)
head(gapminder)
str(gapminder)
```

# Data frames are (special) lists!

```{r}
is.list(gapminder)
length(gapminder)
gapminder[[3]][1:5]
lapply(gapminder, class) 
```

`lapply()` is a function used on lists; it works here to apply the `class()` function to each element of the list, which in this case is each field/column.

# But lists are also vectors!

```{r}
length(gapminder)
someFields <- gapminder[c(3,5)]
head(someFields)
identical(gapminder[c(3,5)], gapminder[ , c(3,5)])
```

In general the placement of commas in R is crucial, but here, two different operations give the same result because of the underlying structure of data frames.

# Factors
- A factor is a special data type in R used for categorical data. In some cases it works like magic and in others it is incredibly frustrating. 

```{r fac}
## let's read the Gapminder data from a file with a special argument:
gapminder <- read.csv(file.path('..', 'data', 'gapminder-FiveYearData.csv'),
          stringsAsFactors = TRUE) # This was the default before R 4.0
class(gapminder$continent)
head(gapminder$continent) # What order are the factors in?
levels(gapminder[["continent"]])  # note alternate way to get the variable
summary(gapminder$continent)
```

- What if we don't like the order these are in? Factor order is important for all kinds of things like plotting, analysis of variance, regression output, and more

# (Advanced) Ordering the Factor
- Ordered factors simply have an additional attribute explaining the order of the levels of a factor
- This is a useful shortcut when we want to preserve some of the meaning provided by the order
- Think ordinal data

This example is a bit artificial as 'continent' doesn't really have a natural ordering.

```{r orderedfac, fig.cap = ""}
gapminder$continent2 <- ordered(gapminder$continent, 
     levels = levels(gapminder$continent)[c(2,1,3,4,5)])

head(gapminder$continent2)
levels(gapminder$continent2)
boxplot(lifeExp ~ continent2, data = gapminder)
```

# (Advanced) Reclassifying Factors
- Turning factors into other data types can be tricky. All factors have an underlying numeric structure.

```{r fac2}
students <- factor(c('basic','proficient','advanced','basic', 
      'advanced', 'minimal'))
levels(students)
unclass(students)
```

- Hmmm, what happened?
- Be careful! The best way to convert a factor is to convert it to a character first.

```{r}
students <- factor(c('basic','proficient','advanced','basic', 
      'advanced', 'minimal'))
score = c(minimal = 3, basic = 1, advanced = 13, proficient = 7) # a named vector
score["advanced"]  # look up by name
students[3]
score[students[3]]
score[as.character(students[3])]
```

What went wrong and how did we fix it?  Notice how easily this could be a big bug in your code.

# Subsetting

There are many ways to select subsets in R. The syntax below is useful for vectors, matrices, data frames, arrays and lists.

```{r}
vec <- gapminder2007$lifeExp
mat <- matrix(1:20, 4, 5)
rownames(mat) <- letters[1:4]
mat
```
1) by direct indexing

```{r}
vec[c(3, 5, 12:14)]
vec[-c(3,5)]
gapminder[c(2,4), 5]
gapminder[c(2,4), 'lifeExp']

## Advanced: subset using a 2-column matrix of indices:
rowInd <- c(1, 3, 4)
colInd <- c(2, 2, 1)
elemInd <- cbind(rowInd, colInd)
elemInd
gapminder[elemInd]
```



2) by a vector of logicals

```{r}
wealthy <- gapminder$gdpPercap > 50000
gapminder$gdpPercap[wealthy]
gapminder[wealthy, ]
```

What happened in the last subsetting operation?

3) by a vector of names
```{r}
mat[c('a', 'd', 'a'), ]
```

4) using *subset()*

```{r}
subset(gapminder, gdpPercap > 50000)
```

5) using *dplyr* tools such as *filter()* and *select()* -- more in Module 6

# Assignment into subsets

We can assign into subsets by using similar syntax, as we saw with vectors.

```{r}
vec <- rnorm(20)
vec[c(3, 5, 12:14)] <- 1:5
vec
mat[2, 3:5] <- rnorm(3)
mat[mat[,1] > 0, ] <- -Inf
mat
```

# Subsetting: quick quiz

**POLL 3D**

*Question 1*: Suppose I want to select the 3rd elements from the 2nd and 4th columns of a matrix or dataframe. Which syntax will work?

Here's a test matrix:

```
mat <- matrix(1:16, nrow = 4, ncol = 4)
```

1) mat[3, (2, 4)]
2) mat[c(FALSE, FALSE, TRUE, FALSE), c(FALSE, TRUE, FALSE, TRUE)]
3) mat[c(FALSE, FALSE, TRUE, FALSE), c(2, 4)]
4) mat[3, c(2, 4)]
5) mat(3, c(2, 4))
6) mat[3, ][c(2, 4)]
7) mat[ , c(2, 4)][3, ]
8) mat[ , c(2, 4)][3]
9) mat[c(2, 4)][3, ]

*Question 2*: (Advanced) One of those answers won't work with a matrix but will work with a dataframe. Which one?

1) mat[3, (2, 4)]
2) mat[c(FALSE, FALSE, TRUE, FALSE), c(FALSE, TRUE, FALSE, TRUE)]
3) mat[c(FALSE, FALSE, TRUE, FALSE), c(2, 4)]
4) mat[3, c(2, 4)]
5) mat(3, c(2, 4))
6) mat[3, ][c(2, 4)]
7) mat[ , c(2, 4)][3, ]
8) mat[ , c(2, 4)][3]
9) mat[c(2, 4)][3, ]


# Strings

R has lots of functionality for character strings. Usually these are stored as vectors of strings, each with arbitrary length.

```{r}
chars <- c('hi', 'hallo', "mother's", 'father\'s', "He said, \"hi\"" )
length(chars)
nchar(chars)
paste("bill", "clinton", sep = " ")  # paste together a set of strings
paste(chars, collapse = ' ')  # paste together things from a vector

strsplit("This is the R bootcamp", split = " ")

countries <- as.character(gapminder2007$country)
substring(countries, 1, 3)
tmp <- countries
substring(tmp, 5, 10) <- "______"
tmp[1:20]
```
We can search for patterns in character vectors and replace patterns (both vectorized!)
```{r}
indexes <- grep("Korea", countries)
indexes
countries[indexes]
countries2 <- gsub("Korea Dem. Rep.", "North Korea", countries)
countries2[indexes]
```

# Regular expressions (regex or regexp)

Some of you may be familiar with using *regular expressions*, which is functionality for doing sophisticated pattern matching and replacement with strings. *Python* and *Perl* are both used extensively for such text manipulation. 

R has a full set of regular expression capabilities available through the *grep()*, *gregexpr()*, and *gsub()* functions (among others - many R functions will work with regular expressions). A particularly nice way to make use of this functionality is to use the *stringr* package, which is more user-friendly than directly using the core R functions.

You can basically do any regular expression/string manipulations in R, though the syntax may be a bit clunky at times.

# More details on reading data into R

Remember that you'll need to know the current working directory so that you know where R is looking for files.

The workhorse for reading into a data frame is *read.table()*, which allows any separator (CSV, tab-delimited, etc.). *read.csv()* is a special case of *read.table()* for CSV files.

You've already seen a bit of this, but let's work through a more involved example, so you can see some of the steps and tricks involved in reading data into R.

```{r, cache=TRUE}
rta <- read.table("../data/RTAData.csv", sep = ",", head = TRUE)
rta[1:5, 1:5]
dim(rta)
# great, we're all set, right?
# Not so fast...
unlist(lapply(rta, class))[1:5]
# let's delve more deeply
# unique(rta[ , 2])  # don't run when creating slides
head(sort(unique(rta[ , 2])))
tail(sort(unique(rta[ , 2])))
# can we handle that with read.table?
# help(read.table)

rta2 <- read.table("../data/RTAData.csv", sep = ",", head = TRUE, 
      na.strings = c('NA', 'x'))
unlist(lapply(rta2, class))[1:5]

# checking...
sum(is.na( rta2[ , 2] ))
sum( rta[ , 2] %in% c('','x'))
```

It's good to first look at your data in plain text format outside of R and then to check it after you've read it into R.

# Other ways to read data into R

The *read.table()* family of functions just skims the surface of things...

1) You can also read in a file as vector of characters, one character string per line of the file with *readLines()*, and then post-process it. 
2) You can read fixed width format (constant number of characters per field) with *read.fwf()*.
3) *read_csv()* (and *read_lines()*, *read_fwf()*, etc.) in the *readr* package is a faster, more helpful drop-in replacement for *read.csv()* that plays well with *dplyr* (see Module 5).
4) the *data.table* package is great for reading and manipulating large datasets (orders of gigabytes or 10s of gigabytes).


# Breakout

### Basics

1) Extract the 5th row from the gapminder dataset.

2) Extract the last row from the gapminder dataset.

3) Count the number of gdpPercap values greater than 50000 in the gapminder dataset.

4) Set all of the gdpPercap values greater than 50000 to NA. You should probably first copy the `gap` object and work on the copy so that the dataset is unchanged (or just read the data into R again afterwards to get a clean copy).

5) Consider the first row of the gapminder dataset, which has Afghanistan for 1952. How do I create a string "Afghanistan-1952" using `gap$country[1]` and `gap$year[1]`? 

### Using the ideas

6) Create a character string using `paste()` that tells the user how many rows there are in the data frame - do this programmatically such that it would work for any data frame regardless of how many rows it has. The result should look like this: "There are 1704 rows in the dataset"

7) If you didn't do it this way already, extract the last row from the gapminder dataset without typing the number '1704'.

8) Create a boolean vector indicating if the life expectancy is greater than 75 and the gdpPercap is less than 10000 and calculate the proportion of all the records these represent.

9) Use that vector to create a new data frame that is a subset of the original data frame.

10) Consider the attributes of the gapminder dataset. What kind of R object is the set of attributes?

### Advanced

11) Create row names for the data frame based on concatenating the Continent, Country, and Year fields.
